---
title: "ST518 Final Project"
author: "Cindy Chenr"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r, message=FALSE, warning=FALSE}
library("MASS")
library("ordinal")
library("nnet")
library("questionr")
library(utils)
library(ggplot2)
library(corrplot)
library(dplyr)
library(tidyverse)
```
Introduction:
Extraction was done by Barry Becker from the 1994 Census database.  A set of 
reasonably clean records was extracted using the following conditions: ((AAGE>16) 
&& (AGI>100) && (AFNLWGT>1)&& (HRSWK>0))

Dataset attributes:

* age: continuous.
* workclass: Private, Self-emp-not-inc, Self-emp-inc, Federal-gov, Local-gov, 
           State-gov, Without-pay, Never-worked.
* fnlwgt: continuous.
* education: Bachelors, Some-college, 11th, HS-grad, Prof-school, Assoc-acdm, 
           Assoc-voc, 9th, 7th-8th, 12th, Masters, 1st-4th, 10th, Doctorate, 
           5th-6th, Preschool.
* education-num: continuous.
* marital-status: Married-civ-spouse, Divorced, Never-married, Separated, Widowed, 
                Married-spouse-absent, Married-AF-spouse.
* occupation: Tech-support, Craft-repair, Other-service, Sales, Exec-managerial, 
            Prof-specialty, Handlers-cleaners, Machine-op-inspct, Adm-clerical, 
            Farming-fishing, Transport-moving, Priv-house-serv, Protective-serv, 
            Armed-Forces.
* relationship: Wife, Own-child, Husband, Not-in-family, Other-relative, Unmarried.
* race: White, Asian-Pac-Islander, Amer-Indian-Eskimo, Other, Black.
* sex: Female, Male.
* capital-gain: continuous.
* capital-loss: continuous.
* hours-per-week: continuous.
* native-country: United-States, Cambodia, England, Puerto-Rico, Canada, Germany, 
                Outlying-US(Guam-USVI-etc), India, Japan, Greece, South, China, 
                Cuba, Iran, Honduras, Philippines, Italy, Poland, Jamaica, Vietnam, 
                Mexico, Portugal, Ireland, France, Dominican-Republic, Laos, 
                Ecuador, Taiwan, Haiti, Columbia, Hungary, Guatemala, Nicaragua, 
                Scotland, Thailand, Yugoslavia, El-Salvador, Trinadad&Tobago, 
                Peru, Hong, Holand-Netherlands.
*income: >50K, <=50K.

The response variable is the `income` which is a Boolean of the class column. 
If the value above 50K per year, then 1 else 0 if they are below 50K per year. 
Then I decide to apply binomial logistical regressions on the datasets.



# Question: Are there any notable patterns or relationships between education and 
income in the dataset?

```{r}

zip_file <- "C:/Users/15306/Desktop/ST 517 Data Analytics II/Final Project/adult.zip"

unzip(zip_file, exdir = tempdir())

data_file <- file.path(tempdir(), "adult.data")

data_lines <- readLines(data_file)

adult_data <- data.frame(do.call(rbind, strsplit(data_lines, ", ", fixed = TRUE)))


str(adult_data)

head(adult_data)

```
```{r}

column_names <- c(
  "age",
  "workclass",
  "fnlwgt",
  "education",
  "education-num",
  "marital-status",
  "occupation",
  "relationship",
  "race",
  "sex",
  "capital-gain",
  "capital-loss",
  "hours-per-week",
  "native-country",
  "income"
)

column_names
```
```{r}
names(adult_data) <- column_names
head(adult_data)
```

```{r}
summary(adult_data)
```
```{r}

if (any(is.na(adult_data))) {
  print("There are missing values in the dataset.")
} else {
  print("There are no missing values in the dataset.")
}


if (any(duplicated(adult_data))) {
  print("There are duplicate rows in the dataset.")
} else {
  print("There are no duplicate rows in the dataset.")
}


duplicate_rows <- adult_data[duplicated(adult_data), ]

adult_data <- adult_data[!duplicated(adult_data), ]
```




```{r}
`marital-status` <- unique(adult_data$`marital-status`)
print(`marital-status` )
```
```{r}

marital_status <- c("Never-married" = 1, "Married-civ-spouse" = 2, "Divorced" = 3, 
          "Married-spouse-absent" = 4, "Separated" = 5, "Married-AF-spouse" = 6, 
                            "Widowed" = 7)


adult_data$marital_status <- marital_status[adult_data$`marital-status`]


head(adult_data$marital_status)

```

```{r}
occupation <- unique(adult_data$occupation)
print(occupation)
```

```{r}
occupation<- c("Adm-clerical" = 1, "Exec-managerial" = 2, "Handlers-cleaners" = 3, 
               "Prof-specialty" = 4, "Other-service" = 5, "Sales" = 6, 
               "Craft-repair" = 7, "Transport-moving" = 8, "Farming-fishing" = 9, 
               "Machine-op-inspct" = 10, "Tech-support" = 11, "?" = 12, 
               "Protective-serv" = 13, "Armed-Forces" = 14, "Priv-house-serv" = 15)


adult_data$occupation <- occupation[adult_data$occupation]

str(adult_data)
```



```{r}
# Convert numeric variables to numeric type
adult_data$age <- as.numeric(adult_data$age)
adult_data$fnlwgt <- as.numeric(adult_data$fnlwgt)
adult_data$`education-num` <- as.numeric(adult_data$`education-num`)
adult_data$`capital-gain` <- as.numeric(adult_data$`capital-gain`)
adult_data$`capital-loss` <- as.numeric(adult_data$`capital-loss`)
adult_data$education <- as.factor(adult_data$education)
adult_data$`hours-per-week` <- as.numeric(adult_data$`hours-per-week`)

str(adult_data)
```


```{r}
numeric_data <- adult_data |> 
  dplyr::select(age, fnlwgt, `education-num`, `capital-gain`, `capital-loss`, 
                `hours-per-week`, marital_status, occupation)

str(numeric_data)

```


```{r}

cor_matrix <- cor(numeric_data)
colnames(cor_matrix) <- c("age", "fnlwgt", "education-num", "capital-gain", 
              "capital-loss", "hours-per-week", "marital_status", "occupation")
rownames(cor_matrix) <- colnames(cor_matrix)

corrplot(cor_matrix, method = "color", type = "full", tl.col = "black", tl.srt = 45)

```


```{r}
describe(numeric_data)
```
```{r}

str(numeric_data)
summary(numeric_data)
any(is.na(numeric_data))

```


```{r}
ggplot(adult_data, aes(x = education, color = income, group = income)) + 
  geom_line(stat = "count") +
  ggtitle("Education by Income") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```
Summary:
From the chart, we could see that for income >50k, the larger number of them have
bachelor, master and some-college degree; for income <50k, the large number of
them has HS-grad or lower, only smaller number of them have bachelor, master and 
some-college degree. We could make assumption on that there is positive relation 
between education and income. However, there is limitation in the analysis like 
the size the dataset. We will need to do further analysis on that.


# Question: How does income vary based on education level, occupation, and other demographic factors?

```{r}
head(adult_data)

adult_data_q2 <- select(adult_data, `education-num`, marital_status, occupation, 
                        `hours-per-week`, age, income)

head(adult_data_q2)

```
```{r}
adult_data_q2$income <- as.factor(adult_data_q2$income)
adult_data_q2$income <- ifelse(adult_data_q2$income == ">50K", 1, 0)

pca_result <- prcomp(adult_data_q2, scale.=TRUE)

# Print PCA summary
summary(pca_result)

```





```{r}

X <- adult_data_q2 %>% select(-income)
y <- adult_data_q2$income

set.seed(123) 
train_index <- sample(nrow(adult_data_q2), 0.8 * nrow(adult_data_q2))
X_train <- X[train_index, ]
y_train <- y[train_index]
X_test <- X[-train_index, ]
y_test <- y[-train_index]



```



```{r}
logit_model <- glm(y_train ~ `education-num` + marital_status + occupation + 
                  `hours-per-week` + age, data = X_train, family = binomial)

predictions <- predict(logit_model, newdata = X_test, type = "response")
predicted_classes <- ifelse(predictions > 0.5, 1, 0)
summary(logit_model)

```


```{r}

confusion_matrix <- table(y_test, predicted_classes)
confusion_matrix
confusion_df <- as.data.frame.matrix(confusion_matrix)
colnames(confusion_df) <- c("0", "1")
confusion_df <- cbind(True_Classes = rownames(confusion_df), confusion_df)
melted_confusion <- reshape2::melt(confusion_df, id.vars = "True_Classes")

heatmap <- ggplot(data = melted_confusion, aes(x = True_Classes, y = variable, 
                                               fill = value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = value), vjust = 1) +
  scale_fill_gradient(low = "white", high = "blue") +  # Use continuous scale
  labs(x = "True Classes", y = "Predicted Classes", title = "Confusion Matrix") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(heatmap)

accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
print(paste("Accuracy:", accuracy))

```
Summary:
This logistic regression model predicts the binary outcome y_train using predictors 
education-num, marital_status, occupation, hours-per-week, and age. Each predictor's 
coefficient indicates its effect on the log odds of the response. The significance 
of each predictor is assessed through the p-values. Lower p-values indicate higher 
significance. All predictors are highly significant and it means they are likely 
important for predicting the response variable. The model fits the data well, 
with a low residual deviance and AIC value. Accuracy: 0.7958  - This metric 
represents the overall correctness of the model's predictions. It's calculated 
as the ratio of the sum of true positives and true negatives to the total number 
of instances. In this case, approximately 79.58% of the predictions are correct.
